---
title: "Range of Motion Analysis"
author: "Jahinna Duplaix"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output:
 html_document:
    code_folding: hide
    toc: true
        toc_depth: 2
    number_sections: true
---

# Project overview

This report presents the final analysis of a range-of-motion dataset collected in leather workers who participated in the 6-month **Ostéart** prevention program.

The research question is:

**What is the effect of the 6-month Ostéart prevention program on wrist extension and forearm supination in leather workers?**

This prevention program was developed in the context of **carpal tunnel syndrome prevention**, as leather workers are highly exposed to repetitive movements and biomechanical constraints affecting the upper limb, especially the wrist and forearm.

The dataset includes **473 employees** measured between **2020 and 2024**.

Two movement variables were measured:
- **wrist extension**
- **forearm supination**

Each participant was measured at two timepoints:
- **T0**: baseline
- **T1**: follow-up after the prevention program

The primary outcomes of interest are the **within-participant changes** between `T0` and `T1`:
- `delta_wrist_extension_deg`
- `delta_forearm_supination_deg`

# Data preparation in Python

The raw dataset was first inspected and prepared in **Python**.

This stage included:
- data cleaning
- harmonization of variable formats
- identification of implausible raw values
- verification of the longitudinal structure
- exclusion of invalid or incomplete observations
- creation of the final complete-case analysis dataset
- computation of the primary outcome variables
- production of descriptive and exploratory outputs

More specifically:
- **3 participants** were excluded because of an invalid longitudinal structure
- **9 participants** were excluded because of incomplete movement data

The final complete-case analysis dataset therefore included **461 participants**.

In Python, the change scores were computed as:

- `delta_wrist_extension_deg = wrist_extension_deg_T1 - wrist_extension_deg_T0`

- `delta_forearm_supination_deg = forearm_supination_deg_T1 - forearm_supination_deg_T0`

Python was also used to produce:

- a descriptive summary table of the main movement variables

- exploratory histograms of the primary outcome variables

The processed dataset exported from Python is used here for the final descriptive analyses, statistical tests, and graphical outputs.

In this report, the final analysis dataset is imported, described, and analysed in order to assess the changes in wrist extension and forearm supination between `T0` and `T1`.

# Setup chunk
```{r setup, include=FALSE}
# Set global chunk options for the whole document
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE)
```

# Load packages
```{r load packages}
# readr is used to import the CSV file exported from Python
library(readr)
# dplyr is used to manipulate and organize data tables
library(dplyr)
# ggplot2 is used to create figures in R
library(ggplot2)
# knitr is used to display cleaner tables in the HTLM report
library(knitr)
```

## Data import and verification

The final complete-case analysis dataset exported from Python was imported into `R`.

This dataset contains one row per participant and includes paired measurements at `T0` and `T1` for wrist extension and forearm supination.

Basic checks were performed to verify:

- dataset dimensions;
- variable names;
- variable types;
- missing values.

## Import the final analysis dataset

```{r import dataset}
#Import the final complete-case analysis dataset exported from Python
analysis_df <- read_csv("data/processed/analysis_dataset.csv")
## First inspection of the dataset
# Display the number of rows and columns
dim(analysis_df)
# Display the names of the variables
names(analysis_df)
# Check dataset structure
glimpse(analysis_df)
# Check missing values by variable
missing_values <- colSums(is.na(analysis_df))
missing_values
```
The imported dataset contained `461` participants and `7` variables.

All expected variables were present: participant identifier, wrist extension at `T0` and `T1`, forearm supination at `T0` and `T1`, and the two change-score variables.

The structure of the dataset was appropriate for the planned analyses. The participant identifier was stored as a character variable, while all movement variables were stored as numeric variables.

No missing values were identified in the final complete-case analysis dataset. Therefore, the dataset exported from Python was considered ready for the statistical analyses in `R`.


## Normality assessment of change scores

The statistical analysis focuses on the change scores between `T0` and `T1`.

Before choosing the paired statistical test, the distribution of the two change scores was assessed. 

This step is necessary because the choice between : 
- a paired t-test; 
- a Wilcoxon signed-rank test 
depends on whether the paired differences are approximately normally distributed.

### Wrist extension change

A Q-Q plot was used to visually assess the distribution of wrist extension change scores.

```{r qqplot-wrist}
# Q-Q plot for wrist extension change
qqnorm(
  analysis_df$delta_wrist_extension_deg,
  main = "Q-Q plot of wrist extension change",
  xlab = "Theoretical quantiles",
  ylab = "Observed quantiles"
)
qqline(analysis_df$delta_wrist_extension_deg)
```

### Forearm supination change

A Q-Q plot was used to visually assess the distribution of forearm supination change scores.

```{r qqplot-supination}
# Q-Q plot for forearm supination change
qqnorm(
  analysis_df$delta_forearm_supination_deg,
  main = "Q-Q plot of forearm supination change",
  xlab = "Theoretical quantiles",
  ylab = "Observed quantiles"
)
qqline(analysis_df$delta_forearm_supination_deg)
```
### Visual interpretation of Q-Q plots

The Q-Q plots show that the central part of both change-score distributions is relatively close to the reference line.

However, deviations are visible at the lower and upper ends of the distributions. This suggests that the change scores are not perfectly normally distributed, especially in the tails.

For wrist extension change, most points follow the reference line in the middle of the distribution, but the extreme values deviate from the line.

For forearm supination change, the same pattern is observed: the central values are close to the line, while the tails show clearer departures from normality.

Therefore, the Q-Q plots suggest that normality should not be assumed without further statistical verification.

```{r shapiro-tests}
# Shapiro-Wilk test for wrist extension change
shapiro.test(analysis_df$delta_wrist_extension_deg)

# Shapiro-Wilk test for forearm supination change
shapiro.test(analysis_df$delta_forearm_supination_deg)
```
