---
title: "Range of Motion Changes After the Ostéart Prevention Program"
author: "Jahinna Duplaix"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
# Restore the R package environment recorded in renv.lock
renv::restore(prompt = FALSE)

# Set global chunk options for the whole document
knitr::opts_chunk$set(
  echo = FALSE, 
  warning = FALSE, 
  message = FALSE)

# Avoid scientific notation in the report
options(scipen = 999)

# readr is used to import the CSV file exported from Python
library(readr)
# dplyr is used to manipulate and organize data tables
library(dplyr) 
# tidyr is used to reshape data for the final figure
library(tidyr)
# ggplot2 is used to create figures in R
library(ggplot2)
# knitr is used to display cleaner tables in the HTML report
library(knitr)
# kableExtra is used to display cleaner tables
library(kableExtra)
```
# Project overview

This report presents the statistical analysis of a range-of-motion **dataset collected between 2023 and 2025 among 473 leather workers** who participated in the 6-month Ostéart prevention program.

The Ostéart program was developed as an occupational prevention program. It focuses on the prevention of upper-limb musculoskeletal disorders, including carpal tunnel syndrome. This context is relevant because leather workers may be exposed to repetitive hand and wrist movements, sustained postures, and biomechanical constraints affecting the wrist, forearm, and hand.

The analysis was based on paired measurements, meaning that the same participants were expected to be measured before and after the 6-month program.

Two range-of-motion outcomes were analysed:

- **wrist extension**, expressed in degrees;
- **forearm supination**, expressed in degrees.

Each participant was measured at two timepoints:

- **T0**: baseline measurement;
- **T1**: follow-up measurement after the 6-month program.

The objective of this analysis was to describe within-participant changes in wrist extension and forearm supination between T0 and T1.

The research question was:

**What changes in wrist extension and forearm supination are observed after the 6-month Ostéart prevention program in leather workers?**

This wording was chosen because the study used a pre-post design without a control group. The analysis can therefore describe within-participant changes between T0 and T1, but it cannot establish a causal effect of the prevention program.

The primary outcome variables were the within-participant change scores:

- `delta_wrist_extension_deg = wrist_extension_deg_T1 - wrist_extension_deg_T0`;
- `delta_forearm_supination_deg = forearm_supination_deg_T1 - forearm_supination_deg_T0`.

Positive delta values indicate an increase in range of motion between T0 and T1. Negative delta values indicate a decrease.


# Data preparation and exploratory analysis in Python

The raw dataset was first inspected and prepared in **Python**.

The Python workflow had four main purposes:

1. to clean and structure the raw dataset;
2. to verify the longitudinal pre-post structure;
3. to identify missing, invalid, or anatomically high values requiring verification;
4. to create the final complete-case dataset and exploratory outputs used in this report.

## Data cleaning and preparation

During data preparation, **3 participants were excluded** because their longitudinal structure was invalid, and **9 participants were excluded** because they had incomplete movement data.

One clearly impossible raw wrist extension value above 180 degrees was identified and recoded as missing...

An anatomical plausibility check was also performed in Python. Seven participants had at least one range-of-motion value above the conservative verification thresholds. These observations were flagged with `anatomical_check_flag`, but they were retained in the main analysis because they were considered high values rather than impossible values.

After these exclusions and checks, the final complete-case analysis dataset included **461 participants**.

Python was also used to compute the two change-score variables defined in the project overview:

- `delta_wrist_extension_deg`;
- `delta_forearm_supination_deg`.

## Descriptive summary of the Python-prepared dataset

Before the inferential analyses in R, descriptive outputs produced in Python were used to examine the prepared dataset.

This descriptive step had two purposes:

1. to summarize the main movement variables at T0 and T1;
2. to visually inspect the distribution of the change scores before choosing the appropriate statistical tests.

### Descriptive table

The descriptive table summarizes the central tendency, dispersion, and range of the two range-of-motion outcomes and their change scores.

It includes:

- the number of complete observations;
- the mean;
- the standard deviation;
- the median;
- the minimum and maximum values.

```{r import-python-descriptive-table}
# Define the path to the descriptive summary table generated in Python
descriptive_summary_file <- "results/tables_exploratory/analysis_summary_table.csv"

# Stop with a clear message if the Python output file is missing
if (!file.exists(descriptive_summary_file)) {
  stop(
    "The descriptive summary table was not found. ",
    "Please run main.ipynb before knitting main.Rmd."
  )
}

# Import the descriptive summary table generated in Python
descriptive_summary <- readr::read_csv(
  descriptive_summary_file,
  show_col_types = FALSE
)

# Display the imported descriptive table
knitr::kable(
  descriptive_summary,
  caption = "Descriptive summary generated in Python",
  align = "lcccccc",
  format = "html"
) %>%  
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "center"
  )
```

The final Python-prepared dataset included **461 participants** with complete paired measurements for both movement outcomes.

For wrist extension, the mean value increased from 50.21 degrees at T0 to 51.89 degrees at T1. The mean change was **1.68** degrees, with a median change of 1.70 degrees.

For forearm supination, the mean value increased from 69.70 degrees at T0 to 71.74 degrees at T1. The mean change was **2.04** degrees, with a median change of 2.10 degrees.

These descriptive results indicate small positive changes between T0 and T1 for both movement outcomes. However, descriptive statistics alone do not indicate whether these changes are statistically significant.

The descriptive summary table was exported to the `results/tables_exploratory` folder as `analysis_summary_table.csv`.

### Histograms of the change-score variables

Histograms were used to visually examine the distribution of the two change-score variables before selecting the appropriate inferential tests.

The two variables displayed were:

- `delta_wrist_extension_deg`;
- `delta_forearm_supination_deg`.

Each change score was computed as `T1 - T0`.

```{r python-histograms, out.width="80%", fig.align="center"}
# Define the paths to the histograms generated in Python
histogram_files <- c(
  "results/figures_exploratory/hist_delta_wrist_extension_deg_global.png",
  "results/figures_exploratory/hist_delta_forearm_supination_deg_global.png"
)

# Stop with a clear message if one or more Python histogram files are missing
if (!all(file.exists(histogram_files))) {
  stop(
    "One or more Python histogram files were not found. ",
    "Please run main.ipynb before knitting main.Rmd."
  )
}

# Insert histograms generated in Python
knitr::include_graphics(histogram_files)
```
The histograms show that the change scores were mainly centered around positive values for both outcomes.

For wrist extension, changes were concentrated around small positive values. This indicates that most participants showed modest increases between T0 and T1.

For forearm supination, changes were also mainly positive, but the distribution was slightly more spread out than for wrist extension.

These exploratory visual results suggest that the two change-score distributions should be examined more formally before choosing the final paired statistical tests.

The next part of the report presents the statistical analysis performed in R. Before assessing normality and running paired tests, the Python-prepared dataset was imported into R and checked to confirm its structure, completeness, and anatomical plausibility flag.

The histograms were exported to the `results/figures_exploratory` folder as:

- `hist_delta_wrist_extension_deg_global.png`; 
- `hist_delta_forearm_supination_deg_global.png`.

# Statistical analysis in R

From this point onward, the analysis was conducted in **R**.

R was used to import the final complete-case dataset exported from Python, verify its structure, assess the distribution of the change scores, perform the paired statistical tests, conduct the sensitivity analysis, and produce the final statistical outputs.

## Data import and verification

The final complete-case analysis dataset exported from Python was imported into R.

This dataset contained one row per participant and included complete paired T0 and T1 measurements for wrist extension and forearm supination.

Before running the inferential analyses, the imported dataset was checked to verify:

- dataset dimensions;
- expected variable names;
- variable types;
- missing values;
- number of anatomically flagged participants.

### Import the final analysis dataset

```{r import-dataset}
# Define the path to the final complete-case dataset exported from Python
analysis_file <- "data/processed/analysis_dataset.csv"

# Stop with a clear message if the Python output file is missing
if (!file.exists(analysis_file)) {
  stop(
    "The final analysis dataset was not found. ",
    "Please run main.ipynb before knitting main.Rmd."
  )
}

# Import the final complete-case analysis dataset exported from Python
analysis_df <- readr::read_csv(
  analysis_file,
  show_col_types = FALSE
)

# Define the variables expected in the imported dataset
expected_variables <- c(
  "participant_id",
  "wrist_extension_deg_T0",
  "wrist_extension_deg_T1",
  "forearm_supination_deg_T0",
  "forearm_supination_deg_T1",
  "anatomical_check_flag",
  "delta_wrist_extension_deg",
  "delta_forearm_supination_deg"
)

# Check that all expected variables are present
missing_expected_variables <- setdiff(expected_variables, names(analysis_df))

if (length(missing_expected_variables) > 0) {
  stop(
    "The following expected variables are missing from analysis_df: ",
    paste(missing_expected_variables, collapse = ", ")
  )
}

# Create a summary of the imported dataset
import_summary <- tibble::tibble(
  Check = c(
    "Number of participants",
    "Number of variables",
    "Total missing values",
    "Number of anatomically flagged participants"
  ),
  Value = c(
    nrow(analysis_df),
    ncol(analysis_df),
    sum(is.na(analysis_df)),
    sum(analysis_df$anatomical_check_flag)
  )
)

# Display the import summary
knitr::kable(
  import_summary,
  caption = "Import summary of the Python-prepared analysis dataset",
  align = "lc",
  format = "html"
) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "center"
  )

# Create a variable-level verification table
variable_summary <- tibble::tibble(
  Variable = names(analysis_df),
  Type = vapply(analysis_df, function(x) class(x)[1], character(1)),
  Missing_values = colSums(is.na(analysis_df))
)

# Display the variable-level verification table
knitr::kable(
  variable_summary,
  caption = "Variable-level verification of the imported dataset",
  align = "llc",
  format = "html"
) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "center"
  )
```

The imported dataset contained **461 participants** and **8 variables**.

All expected variables were present: participant identifier, wrist extension at T0 and T1, forearm supination at T0 and T1, the anatomical plausibility flag, and the two change-score variables.

The structure of the dataset was appropriate for the planned analyses. The participant identifier was stored as a character variable, while all movement variables and change-score variables were stored as numeric variables.

No missing values were identified in the final complete-case analysis dataset. Therefore, the dataset exported from Python was considered ready for the statistical analyses in R.

### Anatomical plausibility flag imported from Python

The anatomical plausibility check was performed in Python before exporting the final analysis dataset.

Participants with values above the conservative anatomical verification thresholds were not excluded from the dataset. They were flagged using the variable `anatomical_check_flag`. This flag identifies participants with high but not physiologically impossible values.

```{r anatomical-flag-summary}
# Count participants according to the anatomical plausibility flag
flag_summary <- analysis_df %>%
  count(anatomical_check_flag, name = "n") %>%
  mutate(
    anatomical_flag_label = ifelse(
      anatomical_check_flag,
      "Flagged for anatomical verification",
      "Not flagged"
    ),
    percentage = round(100 * n / sum(n), 1)
  ) %>%
  select(
    anatomical_flag_label,
    n,
    percentage
  )

# Display a clean summary table
knitr::kable(
  flag_summary,
  caption = "Summary of anatomical plausibility flag",
  col.names = c("Anatomical plausibility flag", "n", "%"),
  align = "lcc",
  format = "html"
) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "center"
  )
```
In the imported R dataset, **7 participants were flagged** and **454 participants were not flagged**.

The flagged participants were retained in the main analysis. They were excluded only later in the sensitivity analysis to verify whether they influenced the main statistical conclusions.

## Normality assessment of change scores

The normality of the two change-score variables was assessed before choosing the appropriate paired statistical test.

Normality was assessed using:

- Visual inspection of Q-Q plots;
- Shapiro-Wilk normality tests.

This step was performed on the final complete-case dataset, including the participants flagged during the anatomical plausibility check. These participants were retained in the main analysis because their values were high but were not considered impossible during data preparation.

### Q-Q plot for wrist extension change

A Q-Q plot was used to visually assess whether wrist extension change followed an approximately normal distribution.

```{r qqplot-wrist, fig.width=7, fig.height=5, fig.align="center"}
# Q-Q plot for wrist extension change
qqnorm(
  analysis_df$delta_wrist_extension_deg,
  main = "Q-Q plot of wrist extension change",
  xlab = "Theoretical quantiles",
  ylab = "Observed quantiles"
)
qqline(analysis_df$delta_wrist_extension_deg)
```

### Q-Q plot for forearm supination change

A Q-Q plot was also used to visually assess whether forearm supination change followed an approximately normal distribution.

```{r qqplot-supination, fig.width=7, fig.height=5, fig.align="center"}
# Q-Q plot for forearm supination change
qqnorm(
  analysis_df$delta_forearm_supination_deg,
  main = "Q-Q plot of forearm supination change",
  xlab = "Theoretical quantiles",
  ylab = "Observed quantiles"
)
qqline(analysis_df$delta_forearm_supination_deg)
```

### Shapiro-Wilk normality tests

Shapiro-Wilk tests were used to statistically assess the normality of the two change-score variables.

The null hypothesis of the Shapiro-Wilk test is that the data are compatible with a normal distribution. A small P value suggests a departure from normality.

```{r shapiro-tests}
# Shapiro-Wilk tests for the two change-score variables
shapiro_wrist <- shapiro.test(analysis_df$delta_wrist_extension_deg)
shapiro_supination <- shapiro.test(analysis_df$delta_forearm_supination_deg)

# Create a clean results table
shapiro_results <- data.frame(
  Outcome = c(
    "Wrist extension change",
    "Forearm supination change"
  ),
  W = round(
    c(
      unname(shapiro_wrist$statistic),
      unname(shapiro_supination$statistic)
    ),
    5
  ),
  p_value_numeric = c(
    shapiro_wrist$p.value,
    shapiro_supination$p.value
  )
)

# Create a formatted P value column for display
shapiro_results$p_value <- format.pval(
  shapiro_results$p_value_numeric,
  digits = 3,
  eps = 0.001
)

# Display the Shapiro-Wilk results
knitr::kable(
  shapiro_results[, c("Outcome", "W", "p_value")],
  caption = "Shapiro-Wilk normality tests for change-score variables",
  col.names = c("Outcome", "W statistic", "P value"),
  align = "lcc",
  format = "html"
) %>%  
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "center"
  )
```
### Normality assessment interpretation

The Q-Q plots showed that most points followed the reference line in the central part of the distribution for both change-score variables. However, deviations were visible at the lower and upper tails.

The Shapiro-Wilk tests were statistically significant for both outcomes.

- For wrist extension change, the test gave **W = `r shapiro_results$W[shapiro_results$Outcome == "Wrist extension change"]` and P = `r shapiro_results$p_value[shapiro_results$Outcome == "Wrist extension change"]`**;
- For forearm supination change, the test gave **W = `r shapiro_results$W[shapiro_results$Outcome == "Forearm supination change"]` and P = `r shapiro_results$p_value[shapiro_results$Outcome == "Forearm supination change"]`**.

Therefore, the assumption of normality was not retained for the two change-score variables. Although the distributions were visually close to normal in their central part, the tail deviations and significant Shapiro-Wilk tests supported the use of a non-parametric paired test.

Because the data are paired measurements from the same participants at T0 and T1, the **Wilcoxon signed-rank test** was selected for the main statistical analysis.

## Paired statistical tests

Because the change-score variables did not meet the normality assumption, paired comparisons between T0 and T1 were performed using the **Wilcoxon signed-rank test**.

The tests were performed separately for wrist extension and forearm supination.

Because the sample size was large and the data contained repeated rounded values, the Wilcoxon tests were performed using the asymptotic approximation rather than the exact calculation.

```{r wilcoxon-tests}
# Wilcoxon signed-rank tests for paired T0-T1 comparisons
wilcox_wrist <- wilcox.test(
  analysis_df$wrist_extension_deg_T1,
  analysis_df$wrist_extension_deg_T0,
  paired = TRUE,
  exact = FALSE,
  conf.int = TRUE
)

wilcox_supination <- wilcox.test(
  analysis_df$forearm_supination_deg_T1,
  analysis_df$forearm_supination_deg_T0,
  paired = TRUE,
  exact = FALSE,
  conf.int = TRUE
)

# Create a clean results table
wilcoxon_results <- data.frame(
  Outcome = c(
    "Wrist extension",
    "Forearm supination"
  ),
  V = c(
    unname(wilcox_wrist$statistic),
    unname(wilcox_supination$statistic)
  ),
  p_value_numeric = c(
    wilcox_wrist$p.value,
    wilcox_supination$p.value
  ),
  Estimated_shift_deg = round(
    c(
      unname(wilcox_wrist$estimate),
      unname(wilcox_supination$estimate)
    ),
    2
  ),
  CI_95 = c(
    paste0(
      sprintf("%.2f", wilcox_wrist$conf.int[1]),
      " to ",
      sprintf("%.2f", wilcox_wrist$conf.int[2])
    ),
    paste0(
      sprintf("%.2f", wilcox_supination$conf.int[1]),
      " to ",
      sprintf("%.2f", wilcox_supination$conf.int[2])
    )
  )
)

# Create a formatted P value column for display
wilcoxon_results$P_value <- format.pval(
  wilcoxon_results$p_value_numeric,
  digits = 3,
  eps = 0.001
)

# Display the table
knitr::kable(
  wilcoxon_results[, c(
    "Outcome",
    "V",
    "P_value",
    "Estimated_shift_deg",
    "CI_95"
  )],
  caption = "Wilcoxon signed-rank tests for paired T0-T1 comparisons",
  col.names = c(
    "Outcome",
    "V statistic",
    "P value",
    "Estimated shift (degrees)",
    "95% confidence interval"
  ),
  align = "lcccc",
  format = "html"
) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "center"
  )
```
### Interpretation of Wilcoxon signed-rank tests

The Wilcoxon signed-rank tests showed statistically significant positive changes between T0 and T1 for both movement variables.

For wrist extension, the test was statistically significant, **V = `r wilcoxon_results$V[wilcoxon_results$Outcome == "Wrist extension"]`, P = `r wilcoxon_results$P_value[wilcoxon_results$Outcome == "Wrist extension"]`**. The estimated location shift was **`r wilcoxon_results$Estimated_shift_deg[wilcoxon_results$Outcome == "Wrist extension"]` degrees**, with a 95% confidence interval from **`r wilcoxon_results$CI_95[wilcoxon_results$Outcome == "Wrist extension"]` degrees**. This indicates a statistically significant increase in wrist extension between T0 and T1.

For forearm supination, the test was also statistically significant, **V = `r wilcoxon_results$V[wilcoxon_results$Outcome == "Forearm supination"]`, P = `r wilcoxon_results$P_value[wilcoxon_results$Outcome == "Forearm supination"]`**. The estimated location shift was **`r wilcoxon_results$Estimated_shift_deg[wilcoxon_results$Outcome == "Forearm supination"]` degrees**, with a 95% confidence interval from **`r wilcoxon_results$CI_95[wilcoxon_results$Outcome == "Forearm supination"]` degrees**. This indicates a statistically significant increase in forearm supination between T0 and T1.

Overall, both wrist extension and forearm supination increased significantly between T0 and T1. However, because this was a pre-post analysis without a control group, these results should be interpreted as observed within-participant changes rather than as definitive causal effects of the prevention program.

## Sensitivity analysis excluding anatomically flagged participants

A sensitivity analysis was performed by excluding participants flagged during the anatomical plausibility check.

The objective was to verify whether the main statistical conclusions were influenced by participants with high anatomical values, including high forearm supination values.

The flagged participants were excluded only for this sensitivity analysis. They were retained in the main analysis.

```{r sensitivity-analysis}
# Create the sensitivity dataset by excluding anatomically flagged participants
analysis_no_flag <- analysis_df %>%
  filter(anatomical_check_flag == FALSE)

# Store the number of participants included and excluded in the sensitivity analysis
n_sensitivity <- nrow(analysis_no_flag)
n_excluded <- nrow(analysis_df) - n_sensitivity

# Wilcoxon signed-rank test for wrist extension, excluding flagged participants
sensitivity_wrist <- wilcox.test(
  analysis_no_flag$wrist_extension_deg_T1,
  analysis_no_flag$wrist_extension_deg_T0,
  paired = TRUE,
  exact = FALSE,
  conf.int = TRUE
)

# Wilcoxon signed-rank test for forearm supination, excluding flagged participants
sensitivity_supination <- wilcox.test(
  analysis_no_flag$forearm_supination_deg_T1,
  analysis_no_flag$forearm_supination_deg_T0,
  paired = TRUE,
  exact = FALSE,
  conf.int = TRUE
)

# Create a clean sensitivity results table
sensitivity_results <- data.frame(
  Outcome = c(
    "Wrist extension",
    "Forearm supination"
  ),
  N = c(
    n_sensitivity,
    n_sensitivity
  ),
  V = c(
    unname(sensitivity_wrist$statistic),
    unname(sensitivity_supination$statistic)
  ),
  p_value_numeric = c(
    sensitivity_wrist$p.value,
    sensitivity_supination$p.value
  ),
  Estimated_shift_deg = round(
    c(
      unname(sensitivity_wrist$estimate),
      unname(sensitivity_supination$estimate)
    ),
    2
  ),
  CI_95 = c(
    paste0(
      sprintf("%.2f", sensitivity_wrist$conf.int[1]),
      " to ",
      sprintf("%.2f", sensitivity_wrist$conf.int[2])
    ),
    paste0(
      sprintf("%.2f", sensitivity_supination$conf.int[1]),
      " to ",
      sprintf("%.2f", sensitivity_supination$conf.int[2])
    )
  )
)

# Create a formatted P value column for display
sensitivity_results$P_value <- format.pval(
  sensitivity_results$p_value_numeric,
  digits = 3,
  eps = 0.001
)

# Display the sensitivity analysis results
knitr::kable(
  sensitivity_results[, c(
    "Outcome",
    "N",
    "V",
    "P_value",
    "Estimated_shift_deg",
    "CI_95"
  )],
  caption = paste0(
    "Sensitivity analysis excluding anatomically flagged participants ",
    "(n excluded = ", n_excluded, ")"
  ),
  col.names = c(
    "Outcome",
    "N",
    "V statistic",
    "P value",
    "Estimated shift (degrees)",
    "95% confidence interval"
  ),
  align = "lccccc",
  format = "html"
) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "center"
  )
```
### Sensitivity analysis interpretation

The sensitivity analysis was performed after excluding the `r n_excluded` participants flagged during the anatomical plausibility check. The sensitivity dataset included `r n_sensitivity` participants.

For wrist extension, the Wilcoxon signed-rank test remained statistically significant, **V = `r sensitivity_results$V[sensitivity_results$Outcome == "Wrist extension"]`, P = `r sensitivity_results$P_value[sensitivity_results$Outcome == "Wrist extension"]`**. The estimated location shift was **`r sensitivity_results$Estimated_shift_deg[sensitivity_results$Outcome == "Wrist extension"]` degrees**, with a 95% confidence interval from **`r sensitivity_results$CI_95[sensitivity_results$Outcome == "Wrist extension"]` degrees**.

For forearm supination, the Wilcoxon signed-rank test also remained statistically significant, **V = `r sensitivity_results$V[sensitivity_results$Outcome == "Forearm supination"]`, P = `r sensitivity_results$P_value[sensitivity_results$Outcome == "Forearm supination"]`**. The estimated location shift was **`r sensitivity_results$Estimated_shift_deg[sensitivity_results$Outcome == "Forearm supination"]` degrees**, with a 95% confidence interval from **`r sensitivity_results$CI_95[sensitivity_results$Outcome == "Forearm supination"]` degrees**.

These results were consistent with the main analysis. Therefore, excluding participants flagged for high anatomical values did not materially change the statistical conclusions.

## Final results tables

The final statistical results are presented in two separate tables for readability.

The first table presents the main analysis including all participants. The second table presents the sensitivity analysis excluding participants flagged during the anatomical plausibility check. 

```{r prepare-final-results-tables}
# Function to format P values consistently
format_p_value <- function(p) {
  ifelse(
    p < 0.001,
    "<0.001",
    sprintf("%.3f", p)
  )
}

# Function to format numeric values with two decimals
format_number <- function(x) {
  sprintf("%.2f", as.numeric(x))
}

# Function to format confidence intervals with two decimals
format_ci <- function(ci) {
  paste0(
    sprintf("%.2f", ci[1]),
    " to ",
    sprintf("%.2f", ci[2])
  )
}

# Create the final results table for the main analysis
final_results_main <- tibble::tibble(
  Outcome = c(
    "Wrist extension",
    "Forearm supination"
  ),
  n = c(
    nrow(analysis_df),
    nrow(analysis_df)
  ),
  `Mean T0 (°)` = c(
    format_number(mean(analysis_df$wrist_extension_deg_T0, na.rm = TRUE)),
    format_number(mean(analysis_df$forearm_supination_deg_T0, na.rm = TRUE))
  ),
  `Mean T1 (°)` = c(
    format_number(mean(analysis_df$wrist_extension_deg_T1, na.rm = TRUE)),
    format_number(mean(analysis_df$forearm_supination_deg_T1, na.rm = TRUE))
  ),
  `Mean delta (°)` = c(
    format_number(mean(analysis_df$delta_wrist_extension_deg, na.rm = TRUE)),
    format_number(mean(analysis_df$delta_forearm_supination_deg, na.rm = TRUE))
  ),
  `Median delta (°)` = c(
    format_number(median(analysis_df$delta_wrist_extension_deg, na.rm = TRUE)),
    format_number(median(analysis_df$delta_forearm_supination_deg, na.rm = TRUE))
  ),
  `Estimated shift (°)` = c(
    format_number(wilcox_wrist$estimate),
    format_number(wilcox_supination$estimate)
  ),
  `95% CI (°)` = c(
    format_ci(wilcox_wrist$conf.int),
    format_ci(wilcox_supination$conf.int)
  ),
  `P value` = c(
    format_p_value(wilcox_wrist$p.value),
    format_p_value(wilcox_supination$p.value)
  )
)

# Create the final results table for the sensitivity analysis
final_results_sensitivity <- tibble::tibble(
  Outcome = c(
    "Wrist extension",
    "Forearm supination"
  ),
  n = c(
    nrow(analysis_no_flag),
    nrow(analysis_no_flag)
  ),
  `Mean T0 (°)` = c(
    format_number(mean(analysis_no_flag$wrist_extension_deg_T0, na.rm = TRUE)),
    format_number(mean(analysis_no_flag$forearm_supination_deg_T0, na.rm = TRUE))
  ),
  `Mean T1 (°)` = c(
    format_number(mean(analysis_no_flag$wrist_extension_deg_T1, na.rm = TRUE)),
    format_number(mean(analysis_no_flag$forearm_supination_deg_T1, na.rm = TRUE))
  ),
  `Mean delta (°)` = c(
    format_number(mean(analysis_no_flag$delta_wrist_extension_deg, na.rm = TRUE)),
    format_number(mean(analysis_no_flag$delta_forearm_supination_deg, na.rm = TRUE))
  ),
  `Median delta (°)` = c(
    format_number(median(analysis_no_flag$delta_wrist_extension_deg, na.rm = TRUE)),
    format_number(median(analysis_no_flag$delta_forearm_supination_deg, na.rm = TRUE))
  ),
  `Estimated shift (°)` = c(
    format_number(sensitivity_wrist$estimate),
    format_number(sensitivity_supination$estimate)
  ),
  `95% CI (°)` = c(
    format_ci(sensitivity_wrist$conf.int),
    format_ci(sensitivity_supination$conf.int)
  ),
  `P value` = c(
    format_p_value(sensitivity_wrist$p.value),
    format_p_value(sensitivity_supination$p.value)
  )
)
```

### Main analysis results table

```{r display-final-results-main}
# Display the final results table for the main analysis
knitr::kable(
  final_results_main,
  caption = "Main analysis results including all participants",
  align = "lcccccccc",
  format = "html"
) %>%
  kableExtra::kable_styling(
    full_width = TRUE,
    position = "center",
    font_size = 13
  ) %>%
  kableExtra::column_spec(
    1,
    width = "4cm",
    extra_css = "white-space: nowrap;"
  ) %>%
  kableExtra::column_spec(
    2:9,
    width = "2.2cm",
    extra_css = "white-space: nowrap; text-align: center;"
  ) %>%
  kableExtra::scroll_box(
    width = "100%"
  )
```

### Sensitivity analysis results table

```{r display-final-results-sensitivity}
# Display the final results table for the sensitivity analysis
knitr::kable(
  final_results_sensitivity,
  caption = "Sensitivity analysis results excluding anatomically flagged participants",
  align = "lcccccccc",
  format = "html"
) %>%
  kableExtra::kable_styling(
    full_width = TRUE,
    position = "center",
    font_size = 13
  ) %>%
  kableExtra::column_spec(
    1,
    width = "4cm",
    extra_css = "white-space: nowrap;"
  ) %>%
  kableExtra::column_spec(
    2:9,
    width = "2.2cm",
    extra_css = "white-space: nowrap; text-align: center;"
  ) %>%
  kableExtra::scroll_box(
    width = "100%"
  )
```

```{r export-final-tables}
# Define the output folder for final results tables
final_tables_dir <- "results/final_tables"

# Create the folder if it does not already exist
dir.create(
  final_tables_dir,
  recursive = TRUE,
  showWarnings = FALSE
)

# Export the final results tables as CSV files
readr::write_csv(
  final_results_main,
  file.path(final_tables_dir, "main_analysis_results.csv")
)

readr::write_csv(
  final_results_sensitivity,
  file.path(final_tables_dir, "sensitivity_analysis_results.csv")
)

# Create HTML versions of the final results tables
main_results_html <- knitr::kable(
  final_results_main,
  caption = "Main analysis results including all participants",
  align = "lcccccccc",
  format = "html"
) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "center"
  )

sensitivity_results_html <- knitr::kable(
  final_results_sensitivity,
  caption = "Sensitivity analysis results excluding anatomically flagged participants",
  align = "lcccccccc",
  format = "html"
) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "center"
  )

# Export the final results tables as HTML files
writeLines(
  as.character(main_results_html),
  file.path(final_tables_dir, "main_analysis_results.html")
)

writeLines(
  as.character(sensitivity_results_html),
  file.path(final_tables_dir, "sensitivity_analysis_results.html")
)
```
The final results tables were exported to the `results/final_tables` folder as:

- `main_analysis_results.csv`;
- `main_analysis_results.html`;
- `sensitivity_analysis_results.csv`;
- `sensitivity_analysis_results.html`.

These files contain the final statistical results for the main analysis and the sensitivity analysis.

```{r extract-final-results-values, include=FALSE}
# Extract rows from the final result tables
main_wrist <- final_results_main[final_results_main$Outcome == "Wrist extension", ]
main_supination <- final_results_main[final_results_main$Outcome == "Forearm supination", ]

sensitivity_wrist_row <- final_results_sensitivity[final_results_sensitivity$Outcome == "Wrist extension", ]
sensitivity_supination_row <- final_results_sensitivity[final_results_sensitivity$Outcome == "Forearm supination", ]

# Main analysis values
main_wrist_mean_t0 <- main_wrist[["Mean T0 (°)"]]
main_wrist_mean_t1 <- main_wrist[["Mean T1 (°)"]]
main_wrist_mean_delta <- main_wrist[["Mean delta (°)"]]
main_wrist_median_delta <- main_wrist[["Median delta (°)"]]
main_wrist_shift <- main_wrist[["Estimated shift (°)"]]
main_wrist_ci <- main_wrist[["95% CI (°)"]]
main_wrist_p <- main_wrist[["P value"]]

main_supination_mean_t0 <- main_supination[["Mean T0 (°)"]]
main_supination_mean_t1 <- main_supination[["Mean T1 (°)"]]
main_supination_mean_delta <- main_supination[["Mean delta (°)"]]
main_supination_median_delta <- main_supination[["Median delta (°)"]]
main_supination_shift <- main_supination[["Estimated shift (°)"]]
main_supination_ci <- main_supination[["95% CI (°)"]]
main_supination_p <- main_supination[["P value"]]

# Sensitivity analysis values
sensitivity_wrist_mean_delta <- sensitivity_wrist_row[["Mean delta (°)"]]
sensitivity_wrist_shift <- sensitivity_wrist_row[["Estimated shift (°)"]]
sensitivity_wrist_ci <- sensitivity_wrist_row[["95% CI (°)"]]

sensitivity_supination_mean_delta <- sensitivity_supination_row[["Mean delta (°)"]]
sensitivity_supination_shift <- sensitivity_supination_row[["Estimated shift (°)"]]
sensitivity_supination_ci <- sensitivity_supination_row[["95% CI (°)"]]
```
### Final results table interpretation

The main analysis included **`r nrow(analysis_df)` participants**. Both wrist extension and forearm supination showed statistically significant positive changes between T0 and T1.

- For wrist extension, the mean value increased from **`r main_wrist_mean_t0` degrees** at T0 to **`r main_wrist_mean_t1` degrees** at T1. The mean delta was **`r main_wrist_mean_delta` degrees** and the median delta was **`r main_wrist_median_delta` degrees**. The Wilcoxon estimated shift was **`r main_wrist_shift` degrees**, with a 95% confidence interval from **`r main_wrist_ci` degrees**. The result was statistically significant (**P = `r main_wrist_p`**).

- For forearm supination, the mean value increased from **`r main_supination_mean_t0` degrees** at T0 to **`r main_supination_mean_t1` degrees** at T1. The mean delta was **`r main_supination_mean_delta` degrees** and the median delta was **`r main_supination_median_delta` degrees**. The Wilcoxon estimated shift was **`r main_supination_shift` degrees**, with a 95% confidence interval from **`r main_supination_ci` degrees**. The result was statistically significant (**P = `r main_supination_p`**).

The sensitivity analysis excluded the **`r n_excluded` participants** flagged during the anatomical plausibility check and included **`r n_sensitivity` participants**. The results remained statistically significant for both outcomes.

- For wrist extension, the mean delta was **`r sensitivity_wrist_mean_delta` degrees** and the Wilcoxon estimated shift was **`r sensitivity_wrist_shift` degrees**, with a 95% confidence interval from **`r sensitivity_wrist_ci` degrees**.

- For forearm supination, the mean delta was **`r sensitivity_supination_mean_delta` degrees** and the Wilcoxon estimated shift was **`r sensitivity_supination_shift` degrees**, with a 95% confidence interval from **`r sensitivity_supination_ci` degrees**.

These sensitivity results were very close to the main analysis results. Therefore, excluding participants flagged for high anatomical values did not materially change the statistical conclusions.

Overall, both wrist extension and forearm supination showed **statistically significant positive changes between T0 and T1**. However, because this was a pre-post analysis without a control group, these findings should be interpreted as observed within-participant changes rather than definitive causal effects of the prevention program.

## Final figures

A final faceted boxplot figure was used to visualize the distributions of wrist extension and forearm supination at T0 and T1.

This figure complements the final statistical tables by showing the distribution of participant values, the median, the interquartile range, and the mean value for each outcome at each timepoint.

The figure summarizes the distributions at T0 and T1. It does not show individual paired trajectories.

```{r final-figure-boxplots, fig.width=9, fig.height=5.5, fig.align="center", fig.cap="Distribution of wrist extension and forearm supination at T0 and T1."}
# Prepare the data for the final figure
final_plot_data <- analysis_df %>%
  select(
    participant_id,
    wrist_extension_deg_T0,
    wrist_extension_deg_T1,
    forearm_supination_deg_T0,
    forearm_supination_deg_T1
  ) %>%
  tidyr::pivot_longer(
    cols = -participant_id,
    names_to = c("Outcome", "Timepoint"),
    names_pattern = "(.*)_deg_(T0|T1)",
    values_to = "Degrees"
  ) %>%
  mutate(
    Outcome = dplyr::recode(
      Outcome,
      "wrist_extension" = "Wrist extension",
      "forearm_supination" = "Forearm supination"
    ),
    Outcome = factor(
      Outcome,
      levels = c("Wrist extension", "Forearm supination")
    ),
    Timepoint = factor(
      Timepoint,
      levels = c("T0", "T1")
    )
  )

# Create the final boxplot figure
final_boxplot <- ggplot(
  final_plot_data,
  aes(x = Timepoint, y = Degrees, fill = Timepoint)
) +
  geom_boxplot(
    alpha = 0.65,
    width = 0.55,
    outlier.shape = NA,
    color = "black"
  ) +
  geom_jitter(
    aes(color = Timepoint),
    position = position_jitter(
      width = 0.12,
      height = 0,
      seed = 123
    ),
    alpha = 0.25,
    size = 1.1
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 23,
    size = 3,
    fill = "white",
    color = "black"
  ) +
  facet_wrap(
    ~ Outcome,
    scales = "free_y"
  ) +
  scale_fill_manual(
    values = c(
      "T0" = "#3B6EA8",
      "T1" = "#C65A1E"
    )
  ) +
  scale_color_manual(
    values = c(
      "T0" = "#3B6EA8",
      "T1" = "#C65A1E"
    )
  ) +
  labs(
    title = "Range-of-motion outcomes at T0 and T1",
    x = "Timepoint",
    y = "Degrees (°)",
    fill = "Timepoint",
    color = "Timepoint"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    strip.text = element_text(face = "bold"),
    legend.position = "bottom"
  )

# Display the figure in the HTML report
final_boxplot

# Define the output folder for final figures
final_figures_dir <- "results/final_figures"

# Create the folder if it does not already exist
dir.create(
  final_figures_dir,
  recursive = TRUE,
  showWarnings = FALSE
)

# Export the final figure
ggsave(
  filename = file.path(final_figures_dir, "final_boxplot_range_of_motion.png"),
  plot = final_boxplot,
  width = 9,
  height = 5.5,
  dpi = 300
)
```
The final boxplot figure was exported to the `results/final_figures` folder as `final_boxplot_range_of_motion.png`.

### Final figure interpretation

The final figure shows the distributions of wrist extension and forearm supination at T0 and T1.

For both outcomes, the T1 distribution appears slightly shifted upward compared with T0. This visual pattern is consistent with the descriptive results and the Wilcoxon signed-rank tests, which showed statistically significant positive changes between T0 and T1.

The individual points show variability between participants. The white diamonds represent the mean values, which were slightly higher at T1 for both wrist extension and forearm supination.

The figure should be interpreted as a descriptive visual summary of the observed pre-post changes. It is visually consistent with the statistical results, but it does not establish a causal effect of the Ostéart prevention program because the study design did not include a control group.

# Discussion

The aim of this project was to describe changes in wrist extension and forearm supination between T0 and T1 in leather workers who participated in the 6-month Ostéart prevention program.

The main analysis showed statistically significant positive changes for both outcomes. Wrist extension increased by a mean of **1.68 degrees**, and forearm supination increased by a mean of **2.04 degrees**. The Wilcoxon estimated shifts were **1.70 degrees** for wrist extension and **2.10 degrees** for forearm supination. These results indicate a positive change between baseline and follow-up for both range-of-motion outcomes.

However, the magnitude of the observed changes was modest. This distinction is important because statistical significance does not necessarily imply a large clinical or functional change. With a sample of **461 participants**, even small changes can reach statistical significance. Therefore, the results should be interpreted as statistically significant and consistent in the sensitivity analysis, but modest in size.

The observed range-of-motion values were compared with reference values reported by Norkin and White. For wrist extension, the AAOS reference value is **70 degrees**, while the AMA considers **60 degrees or more** as functional. In the present dataset, mean wrist extension increased from **50.21 degrees at T0** to **51.89 degrees at T1**. Therefore, although the change was statistically significant, the mean values remained below these reference values.

For forearm supination, the AAOS reference value is **80 degrees**, while the AMA considers **80 degrees or more** as functional. In the present dataset, mean forearm supination increased from **69.70 degrees at T0** to **71.74 degrees at T1**. These mean values also remained below the AAOS and AMA reference values. This suggests that the observed increase did not bring the group mean up to usual reference values, but it did indicate a small positive shift.

These comparisons suggest that the observed mean values were anatomically plausible but lower than commonly reported reference values. However, reference values should be interpreted cautiously. Range-of-motion values can vary according to age, sex, limb dominance, measurement method, occupational exposure, pain, fatigue, and assessment conditions. In this project, the participants were leather workers, a population potentially exposed to repetitive hand and wrist movements, sustained postures, and biomechanical constraints affecting the wrist, forearm, and hand. This occupational context may partly explain lower mean values than those reported in general reference sources.

The sensitivity analysis excluding the **7 participants** flagged during the anatomical plausibility check produced results very close to the main analysis. This suggests that excluding participants flagged for high anatomical values did not materially change the statistical conclusions. This also supports the decision to retain them in the main analysis while checking their influence separately.

Direct comparison of the observed deltas with previous intervention studies is limited. No directly comparable study was identified during this project involving leather workers, a 6-month prevention program, and the same two range-of-motion outcomes. Therefore, the results should mainly be interpreted as observed within-participant changes in this dataset rather than as effects directly comparable to previous intervention trials.

Finally, because the study used a pre-post design without a control group, the results cannot establish a causal effect of the Ostéart prevention program. They indicate modest observed improvements over time, but other explanations cannot be ruled out, such as measurement variability, natural variation, changes in work exposure, or other contextual factors during the follow-up period.

# Limitations

This analysis has several limitations.

First, the study used a pre-post design without a control group. Therefore, the results can describe observed within-participant changes between T0 and T1, but they cannot establish a causal effect of the Ostéart prevention program.

Second, the analysis was limited to participants with complete paired data for both movement outcomes. Participants with incomplete T0 or T1 movement data were excluded from the final complete-case dataset. This complete-case approach was necessary for the paired analysis, but it may have excluded some participants from the original dataset.

Third, some high anatomical values were identified during data preparation. These values were not removed from the main analysis because they were not considered impossible, but they were flagged and assessed through a sensitivity analysis.

Fourth, statistically significant changes do not necessarily imply clinically meaningful changes. The observed mean changes were small in magnitude and should therefore be interpreted cautiously.

Fifth, this analysis focused only on wrist extension and forearm supination. It did not assess other clinical outcomes related to upper-limb musculoskeletal disorders or carpal tunnel syndrome, such as pain, symptoms, functional limitation, grip strength, or work-related disability.

Finally, range-of-motion measurements may be influenced by measurement conditions, examiner variability, participant fatigue, pain, motivation, and occupational exposure during the follow-up period. These factors could not be fully controlled in this analysis.

# Conclusion

This project analysed range-of-motion data collected between 2023 and 2025 among leather workers who participated in the 6-month Ostéart prevention program.

After Python data preparation, the final complete-case analysis dataset included **461 participants**. The statistical analysis was then performed in R.

Both wrist extension and forearm supination showed statistically significant positive changes between T0 and T1. Wrist extension increased by a mean of **1.68 degrees**, and forearm supination increased by a mean of **2.04 degrees**. The Wilcoxon signed-rank tests confirmed statistically significant increases for both outcomes.

The sensitivity analysis excluding the **7 participants** flagged during the anatomical plausibility check produced results consistent with the main analysis. Excluding these participants did not materially change the statistical conclusions.

Overall, the findings suggest modest observed improvements in wrist extension and forearm supination between baseline and follow-up. However, because the study used a pre-post design without a control group, these findings should be interpreted as observed within-participant changes rather than definitive causal effects of the Ostéart prevention program.

# References

Norkin, C. C., & White, D. J. (2016). *Measurement of Joint Motion: A Guide to Goniometry* (5th ed.). F. A. Davis Company.

TM Institute. (n.d.). *Ostéart Pro*. Accessed May 11, 2026. https://www.tminstitute.fr/osteart.html

Palmer, K. T., Harris, E. C., & Coggon, D. (2007). Carpal tunnel syndrome and its relation to occupation: A systematic literature review. *Occupational Medicine, 57*(1), 57–66. doi:10.1093/occmed/kql125

Nicoletti, S., Carino, M., Di Leone, G., et al. (2008). Prevalence of upper limb work-related musculoskeletal disorders (UL-WMSDs) in workers of the upholstered furniture industry. *La Medicina del Lavoro, 99*(4), 271–280. doi:10.23749/mdl.v99i4.1683