---
title: "Range of Motion Changes After the Ostéart Prevention Program"
author: "Jahinna Duplaix"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---
```{r setup, include=FALSE}
# Set global chunk options for the whole document
knitr::opts_chunk$set(
  echo = FALSE, 
  warning = FALSE, 
  message = FALSE)

# readr is used to import the CSV file exported from Python
library(readr)
# dplyr is used to manipulate and organize data tables
library(dplyr) 
# ggplot2 is used to create figures in R
library(ggplot2)
# knitr is used to display cleaner tables in the HTML report
library(knitr)
# kableExtra is used to display cleaner tables
library(kableExtra)
```
# Project overview

This report presents the statistical analysis of a range-of-motion **dataset collected between 2023 and 2025 among 473 leather workers** who participated in the 6-month Ostéart prevention program.

The Ostéart program was developed as an occupational prevention program. It focuses on the prevention of upper-limb musculoskeletal disorders, including carpal tunnel syndrome. This context is relevant because leather workers may be exposed to repetitive hand and wrist movements, sustained postures, and biomechanical constraints affecting the wrist, forearm, and hand.

The dataset contained paired measurements, meaning that the same participants were measured before and after the 6-month program.

Two range-of-motion outcomes were analysed:

- **wrist extension**, expressed in degrees;
- **forearm supination**, expressed in degrees.

Each participant was measured at two timepoints:

- **T0**: baseline measurement;
- **T1**: follow-up measurement after the 6-month program.

The objective of this analysis was to assess within-participant changes in wrist extension and forearm supination between T0 and T1.

The research question was:

**What changes in wrist extension and forearm supination are observed after the 6-month Ostéart prevention program in leather workers?**

This wording was chosen because the study used a pre-post design without a control group. The analysis can therefore describe within-participant changes between T0 and T1, but it cannot establish a causal effect of the prevention program.

The primary outcome variables were the within-participant change scores:

- `delta_wrist_extension_deg = wrist_extension_deg_T1 - wrist_extension_deg_T0`;
- `delta_forearm_supination_deg = forearm_supination_deg_T1 - forearm_supination_deg_T0`.

Positive delta values indicate an increase in range of motion between T0 and T1. Negative delta values indicate a decrease.


# Data preparation and exploratory analysis in Python

The raw dataset was first inspected and prepared in **Python**.

The Python workflow had four main purposes:

1. to clean and structure the raw dataset;
2. to verify the longitudinal pre-post structure;
3. to identify missing, invalid, or anatomically questionable values;
4. to create the final complete-case dataset and exploratory outputs used in this report.

## Data cleaning and preparation

During data preparation, **3 participants were excluded** because their longitudinal structure was invalid, and **9 participants were excluded** because they had incomplete movement data.

One clearly impossible raw value for wrist extension was identified and recoded as missing before the complete-case dataset was created.

An anatomical plausibility check was also performed in Python. Seven participants had forearm supination values above the conservative verification threshold of 90 degrees. These observations were flagged with `anatomical_check_flag`, but they were retained in the main analysis because they were considered high values rather than impossible values.

After these exclusions and checks, the final complete-case analysis dataset included **461 participants**.

Python was also used to compute the two change-score variables defined in the project overview:

- `delta_wrist_extension_deg`;
- `delta_forearm_supination_deg`.

## Descriptive summary of the Python-prepared dataset

Before the inferential analyses in R, descriptive outputs produced in Python were used to examine the prepared dataset.

This descriptive step had two purposes:

1. to summarize the main movement variables at T0 and T1;
2. to visually inspect the distribution of the change scores before choosing the appropriate statistical tests.

### Descriptive table

The descriptive table summarizes the central tendency, dispersion, and range of the two range-of-motion outcomes and their change scores.

It includes:

- the number of complete observations;
- the mean;
- the standard deviation;
- the median;
- the minimum and maximum values.

```{r import-python-descriptive-table}
# Import descriptive summary table generated in Python
descriptive_summary <- readr::read_csv(
  "results/tables_exploratory/analysis_summary_table.csv",
  show_col_types = FALSE
)

# Display the imported descriptive table
knitr::kable(
  descriptive_summary,
  caption = "Descriptive summary generated in Python",
  align = "lcccccc",
  format = "html"
) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "center"
  )
```

The final Python-prepared dataset included **461 participants** with complete paired measurements.

For wrist extension, the mean value increased from 50.21 degrees at T0 to 51.89 degrees at T1. The mean change was **1.68** degrees, with a median change of 1.70 degrees.

For forearm supination, the mean value increased from 69.70 degrees at T0 to 71.74 degrees at T1. The mean change was **2.04** degrees, with a median change of 2.10 degrees.

These descriptive results indicate small positive changes between T0 and T1 for both movement outcomes. However, descriptive statistics alone do not indicate whether these changes are statistically significant.

The descriptive summary table was exported to the `results/tables_exploratory` folder as `analysis_summary_table.csv`

### Histograms of the change-score variables

Histograms were used to visually examine the distribution of the two change-score variables before the normality assessment.

The two variables displayed were:

- `delta_wrist_extension_deg`;
- `delta_forearm_supination_deg`.

Each change score was computed as `T1 - T0`.

```{r python-histograms, out.width="80%", fig.align="center"}
# Insert histograms generated in Python
knitr::include_graphics(c(
  "results/figures_exploratory/hist_delta_wrist_extension_deg_global.png",
  "results/figures_exploratory/hist_delta_forearm_supination_deg_global.png"
))
```
The histograms show that most change scores were positive for both outcomes.

For wrist extension, changes were concentrated around small positive values. This indicates that most participants showed modest increases between T0 and T1.

For forearm supination, changes were also mainly positive, but the distribution was slightly more spread out than for wrist extension.

These exploratory visual results suggest that the two change-score distributions should be examined more formally before choosing the final paired statistical tests.

The next part of the report presents the statistical analysis performed in R. Before assessing normality and running paired tests, the Python-prepared dataset was imported into R and checked to confirm its structure, completeness, and anatomical plausibility flag.

The histograms were exported to the `results/figures_exploratory` folder as:

- `hist_delta_wrist_extension_deg_global.png`; 
- `hist_delta_forearm_supination_deg_global.png`

# Statistical analysis in R

From this point onward, the analysis was conducted in **R**. 

R was used to import the final complete-case dataset exported from Python, verify its structure, assess the distribution of the change scores, perform the paired statistical tests, conduct the sensitivity analysis, and produce the final statistical outputs.

Before running the inferential analyses, the imported dataset was checked to ensure that it was complete, correctly structured, and consistent with the Python-prepared analysis dataset.

## Data import and verification

The final complete-case analysis dataset exported from Python was imported into R.

This dataset contained one row per participant and included paired T0 and T1 measurements for wrist extension and forearm supination.

The import verification checked:

- dataset dimensions;
- expected variable names;
- variable types;
- missing values;
- number of anatomically flagged participants.

### Import the final analysis dataset

```{r import-dataset}
# Import the final complete-case analysis dataset exported from Python
analysis_df <- read_csv(
  "data/processed/analysis_dataset.csv",
  show_col_types = FALSE
)

# Basic import checks
dim(analysis_df)
names(analysis_df)
glimpse(analysis_df)
colSums(is.na(analysis_df))
```

The imported dataset contained **461 participants** and **8 variables**.

All expected variables were present: participant identifier, wrist extension at T0 and T1, forearm supination at T0 and T1, the anatomical plausibility flag, and the two change-score variables.

The structure of the dataset was appropriate for the planned analyses. The participant identifier was stored as a character variable, while all movement variables and change-score variables were stored as numeric variables.

No missing values were identified in the final complete-case analysis dataset. Therefore, the dataset exported from Python was considered ready for the statistical analyses in R.

### Anatomical plausibility flag imported from Python

The anatomical plausibility check was performed in Python before exporting the final analysis dataset.

Participants with values above the conservative anatomical verification thresholds were not excluded from the dataset. They were flagged using the variable `anatomical_check_flag`. This flag identifies participants with high but not physiologically impossible values.

```{r anatomical-flag-summary}
# Count participants according to the anatomical plausibility flag
flag_summary <- analysis_df %>%
  count(anatomical_check_flag, name = "n") %>%
  mutate(
    percentage = round(100 * n / sum(n), 1)
  )

# Display a clean summary table
knitr::kable(
  flag_summary,
  caption = "Summary of anatomical plausibility flag",
  col.names = c("Anatomical plausibility flag", "n", "%"),
  align = "lcc",
  format = "html"
) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "center"
  )
```

```{r anatomical-flag-participants-check}
# Display participants flagged during the anatomical plausibility check
flagged_participants <- analysis_df %>%
  filter(anatomical_check_flag == TRUE) %>%
  select(
    participant_id,
    wrist_extension_deg_T0,
    wrist_extension_deg_T1,
    forearm_supination_deg_T0,
    forearm_supination_deg_T1,
    anatomical_check_flag
  )

knitr::kable(
  flagged_participants,
  caption = "Participants flagged during anatomical plausibility check",
  align = "lrrrrl",
  format = "html"
) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "center"
  )
```
In the imported R dataset, **7 participants were flagged** and **454 participants were not flagged**.

The flagged participants were retained in the main analysis. They were excluded only later in the sensitivity analysis to verify whether they influenced the main statistical conclusions.

## Normality assessment of change scores

The normality of the two change-score variables was assessed before choosing the appropriate paired statistical test.

Normality was assessed using:

- Visual inspection of Q-Q plots;
- Shapiro-Wilk normality tests.

This step was performed on the final complete-case dataset, including the participants flagged during the anatomical plausibility check. These participants were retained in the main analysis because their values were high but not physiologically impossible.

### Q-Q plot for wrist extension change

A Q-Q plot was used to visually assess whether wrist extension change followed an approximately normal distribution.

```{r qqplot-wrist, fig.width=7, fig.height=5, fig.align="center"}
# Q-Q plot for wrist extension change
qqnorm(
  analysis_df$delta_wrist_extension_deg,
  main = "Q-Q plot of wrist extension change",
  xlab = "Theoretical quantiles",
  ylab = "Observed quantiles"
)
qqline(analysis_df$delta_wrist_extension_deg)
```

### Q-Q plot for forearm supination change

A Q-Q plot was also used to visually assess whether forearm supination change followed an approximately normal distribution.

```{r qqplot-supination, fig.width=7, fig.height=5, fig.align="center"}
# Q-Q plot for forearm supination change
qqnorm(
  analysis_df$delta_forearm_supination_deg,
  main = "Q-Q plot of forearm supination change",
  xlab = "Theoretical quantiles",
  ylab = "Observed quantiles"
)
qqline(analysis_df$delta_forearm_supination_deg)
```

### Shapiro-Wilk normality tests

Shapiro-Wilk tests were used to statistically assess the normality of the two change-score variables.

The null hypothesis of the Shapiro-Wilk test is that the data are compatible with a normal distribution.

```{r shapiro-tests}
# Shapiro-Wilk tests for the two change-score variables
shapiro_wrist <- shapiro.test(analysis_df$delta_wrist_extension_deg)
shapiro_supination <- shapiro.test(analysis_df$delta_forearm_supination_deg)

# Create a clean results table
shapiro_results <- data.frame(
  Outcome = c(
    "Wrist extension change",
    "Forearm supination change"
  ),
  W = round(
    c(
      unname(shapiro_wrist$statistic),
      unname(shapiro_supination$statistic)
    ),
    5
  ),
  p_value = c(
    format.pval(shapiro_wrist$p.value, digits = 3, eps = 0.001),
    format.pval(shapiro_supination$p.value, digits = 3, eps = 0.001)
  )
)

# Display the Shapiro-Wilk results
knitr::kable(
  shapiro_results,
  caption = "Shapiro-Wilk normality tests for change-score variables",
  col.names = c("Outcome", "W statistic", "P value"),
  align = "lcc",
  format = "html"
) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "center"
  )
```
### Normality assessment interpretation

The Q-Q plots showed that most points followed the reference line in the central part of the distribution for both change-score variables. However, **deviations were visible at the lower and upper tails.**

The Shapiro-Wilk tests were statistically significant for both outcomes. 

- For wrist extension change, the test gave **W = 0.99132 and P = 0.008421**;
- For forearm supination change, the test gave **W = 0.99058 and P = 0.004832**.

Therefore, the assumption of normality was not retained for the two change-score variables. Although the distributions were visually close to normal in their central part, the tail deviations and significant Shapiro-Wilk tests supported the use of a non-parametric paired test.

Because the data are paired measurements from the same participants at T0 and T1, the **Wilcoxon signed-rank test** was selected for the main statistical analysis.

## Paired statistical tests

Because the change-score variables did not meet the normality assumption, paired comparisons between T0 and T1 were performed using the **Wilcoxon signed-rank test**.

The tests were performed separately for wrist extension and forearm supination.

Because the sample size was large and the data contained repeated rounded values, the Wilcoxon tests were performed using the asymptotic approximation rather than the exact calculation.

```{r wilcoxon-tests}
# Wilcoxon signed-rank tests for paired T0-T1 comparisons
wilcox_wrist <- wilcox.test(
  analysis_df$wrist_extension_deg_T1,
  analysis_df$wrist_extension_deg_T0,
  paired = TRUE,
  exact = FALSE,
  conf.int = TRUE
)

wilcox_supination <- wilcox.test(
  analysis_df$forearm_supination_deg_T1,
  analysis_df$forearm_supination_deg_T0,
  paired = TRUE,
  exact = FALSE,
  conf.int = TRUE
)

# Create a clean results table
wilcoxon_results <- data.frame(
  Outcome = c(
    "Wrist extension",
    "Forearm supination"
  ),
  V = c(
    unname(wilcox_wrist$statistic),
    unname(wilcox_supination$statistic)
  ),
  P_value = c(
    format.pval(wilcox_wrist$p.value, digits = 3, eps = 0.001),
    format.pval(wilcox_supination$p.value, digits = 3, eps = 0.001)
  ),
  Estimated_shift_deg = sprintf(
    "%.2f",
    c(
      unname(wilcox_wrist$estimate),
      unname(wilcox_supination$estimate)
    )
  ),
  CI_95 = c(
    paste0(
      sprintf("%.2f", wilcox_wrist$conf.int[1]),
      " to ",
      sprintf("%.2f", wilcox_wrist$conf.int[2])
    ),
    paste0(
      sprintf("%.2f", wilcox_supination$conf.int[1]),
      " to ",
      sprintf("%.2f", wilcox_supination$conf.int[2])
    )
  )
)

# Display the table
knitr::kable(
  wilcoxon_results,
  caption = "Wilcoxon signed-rank tests for paired T0-T1 comparisons",
  col.names = c(
    "Outcome",
    "V statistic",
    "P value",
    "Estimated shift (degrees)",
    "95% confidence interval"
  ),
  align = "lcccc",
  format = "html"
) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "center"
  )
```
_Interpretation of Wilcoxon signed-rank tests_:

The Wilcoxon signed-rank tests showed statistically significant positive changes between T0 and T1 for both movement variables.

For wrist extension, the test was statistically significant, **V = 100542, P < 0.001**. The estimated location shift was **1.70 degrees**, with a 95% confidence interval from **1.60 to 1.85 degrees**. This indicates a statistically significant increase in wrist extension between T0 and T1.

For forearm supination, the test was also statistically significant, **V = 99224, P < 0.001**. The estimated location shift was **2.10 degrees**, with a 95% confidence interval from **1.90 to 2.25 degrees**. This indicates a statistically significant increase in forearm supination between T0 and T1.

Overall, both wrist extension and forearm supination increased significantly between T0 and T1. However, because this was a pre-post analysis without a control group, these results should be interpreted as observed within-participant changes rather than as definitive causal effects of the prevention program.

## Sensitivity analysis excluding anatomically flagged participants

A sensitivity analysis was performed by excluding participants flagged during the anatomical plausibility check.

The objective was to verify whether the main statistical conclusions were influenced by participants with high anatomical values, especially high forearm supination values.

The flagged participants were excluded only for this sensitivity analysis. They were retained in the main analysis.

```{r sensitivity-analysis}
# Create the sensitivity dataset by excluding anatomically flagged participants
analysis_no_flag <- analysis_df %>%
  filter(anatomical_check_flag == FALSE)

# Store the number of participants included in the sensitivity analysis
n_sensitivity <- nrow(analysis_no_flag)
n_excluded <- nrow(analysis_df) - n_sensitivity

# Wilcoxon signed-rank test for wrist extension, excluding flagged participants
sensitivity_wrist <- wilcox.test(
  analysis_no_flag$wrist_extension_deg_T1,
  analysis_no_flag$wrist_extension_deg_T0,
  paired = TRUE,
  exact = FALSE,
  conf.int = TRUE
)

# Wilcoxon signed-rank test for forearm supination, excluding flagged participants
sensitivity_supination <- wilcox.test(
  analysis_no_flag$forearm_supination_deg_T1,
  analysis_no_flag$forearm_supination_deg_T0,
  paired = TRUE,
  exact = FALSE,
  conf.int = TRUE
)

# Create a clean sensitivity results table
sensitivity_results <- data.frame(
  Outcome = c(
    "Wrist extension",
    "Forearm supination"
  ),
  N = c(
    n_sensitivity,
    n_sensitivity
  ),
  V = c(
    unname(sensitivity_wrist$statistic),
    unname(sensitivity_supination$statistic)
  ),
  P_value = c(
    format.pval(sensitivity_wrist$p.value, digits = 3, eps = 0.001),
    format.pval(sensitivity_supination$p.value, digits = 3, eps = 0.001)
  ),
  Estimated_shift_deg = round(
    c(
      unname(sensitivity_wrist$estimate),
      unname(sensitivity_supination$estimate)
    ),
    2
  ),
  CI_95 = c(
    paste0(
      sprintf("%.2f", sensitivity_wrist$conf.int[1]),
      " to ",
      sprintf("%.2f", sensitivity_wrist$conf.int[2])
    ),
    paste0(
      sprintf("%.2f", sensitivity_supination$conf.int[1]),
      " to ",
  sprintf("%.2f", sensitivity_supination$conf.int[2])
    )
  )
)

# Display the sensitivity analysis results
knitr::kable(
  sensitivity_results,
  caption = "Sensitivity analysis excluding anatomically flagged participants",
  col.names = c(
    "Outcome",
    "N",
    "V statistic",
    "P value",
    "Estimated shift (degrees)",
    "95% confidence interval"
  ),
  align = "lccccc",
  format = "html"
) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "center"
  )
```
_Sensitivity analysis interpretation_:

The sensitivity analysis was performed **after excluding the 7 participants flagged during the anatomical plausibility check**. The sensitivity dataset included **454 participants**.

For wrist extension, the Wilcoxon signed-rank test remained statistically significant, **V = 97504, P < 0.001**. The estimated location shift was **1.70 degrees**, with a 95% confidence interval from **1.60 to 1.85 degrees**.

For forearm supination, the Wilcoxon signed-rank test also remained statistically significant, **V = 96135, P < 0.001**. The estimated location shift was **2.05 degrees**, with a 95% confidence interval from **1.90 to 2.25 degrees**.

These results were consistent with the main analysis. Therefore, the **observed increases in wrist extension and forearm supination were not driven by the participants flagged for high anatomical values.**

## Final results tables

The final statistical results are presented in two separate tables for readability.

The first table presents the main analysis including all participants. The second table presents the sensitivity analysis excluding participants flagged during the anatomical plausibility check. 

```{r prepare-final-results-tables}
# Function to format P values consistently
format_p_value <- function(p) {
  ifelse(
    p < 0.001,
    "<0.001",
    sprintf("%.3f", p)
  )
}

# Function to format numeric values with two decimals
format_number <- function(x) {
  sprintf("%.2f", as.numeric(x))
}

# Function to format confidence intervals with two decimals
format_ci <- function(ci) {
  paste0(
    sprintf("%.2f", ci[1]),
    " to ",
    sprintf("%.2f", ci[2])
  )
}

# Create the final results table for the main analysis
final_results_main <- tibble::tibble(
  Outcome = c(
    "Wrist extension",
    "Forearm supination"
  ),
  n = c(
    nrow(analysis_df),
    nrow(analysis_df)
  ),
  `Mean T0 (°)` = c(
    format_number(mean(analysis_df$wrist_extension_deg_T0, na.rm = TRUE)),
    format_number(mean(analysis_df$forearm_supination_deg_T0, na.rm = TRUE))
  ),
  `Mean T1 (°)` = c(
    format_number(mean(analysis_df$wrist_extension_deg_T1, na.rm = TRUE)),
    format_number(mean(analysis_df$forearm_supination_deg_T1, na.rm = TRUE))
  ),
  `Mean delta (°)` = c(
    format_number(mean(analysis_df$delta_wrist_extension_deg, na.rm = TRUE)),
    format_number(mean(analysis_df$delta_forearm_supination_deg, na.rm = TRUE))
  ),
  `Median delta (°)` = c(
    format_number(median(analysis_df$delta_wrist_extension_deg, na.rm = TRUE)),
    format_number(median(analysis_df$delta_forearm_supination_deg, na.rm = TRUE))
  ),
  `Estimated shift (°)` = c(
    format_number(wilcox_wrist$estimate),
    format_number(wilcox_supination$estimate)
  ),
  `95% CI (°)` = c(
    format_ci(wilcox_wrist$conf.int),
    format_ci(wilcox_supination$conf.int)
  ),
  `P value` = c(
    format_p_value(wilcox_wrist$p.value),
    format_p_value(wilcox_supination$p.value)
  )
)
# Create the final results table for the sensitivity analysis
final_results_sensitivity <- tibble::tibble(
  Outcome = c(
    "Wrist extension",
    "Forearm supination"
  ),
  n = c(
    nrow(analysis_no_flag),
    nrow(analysis_no_flag)
  ),
  `Mean T0 (°)` = c(
    format_number(mean(analysis_no_flag$wrist_extension_deg_T0, na.rm = TRUE)),
    format_number(mean(analysis_no_flag$forearm_supination_deg_T0, na.rm = TRUE))
  ),
  `Mean T1 (°)` = c(
    format_number(mean(analysis_no_flag$wrist_extension_deg_T1, na.rm = TRUE)),
    format_number(mean(analysis_no_flag$forearm_supination_deg_T1, na.rm = TRUE))
  ),
  `Mean delta (°)` = c(
    format_number(mean(analysis_no_flag$delta_wrist_extension_deg, na.rm = TRUE)),
    format_number(mean(analysis_no_flag$delta_forearm_supination_deg, na.rm = TRUE))
  ),
  `Median delta (°)` = c(
    format_number(median(analysis_no_flag$delta_wrist_extension_deg, na.rm = TRUE)),
    format_number(median(analysis_no_flag$delta_forearm_supination_deg, na.rm = TRUE))
  ),
  `Estimated shift (°)` = c(
    format_number(sensitivity_wrist$estimate),
    format_number(sensitivity_supination$estimate)
  ),
  `95% CI (°)` = c(
    format_ci(sensitivity_wrist$conf.int),
    format_ci(sensitivity_supination$conf.int)
  ),
  `P value` = c(
    format_p_value(sensitivity_wrist$p.value),
    format_p_value(sensitivity_supination$p.value)
  )
)

```

### Main analysis results table

```{r display-final-results-main}
# Display the final results table for the main analysis
knitr::kable(
  final_results_main,
  caption = "Main analysis results",
  align = "lcccccccc",
  format = "html"
) %>%
  kableExtra::kable_styling(
    full_width = TRUE,
    position = "center",
    font_size = 13
  ) %>%
  kableExtra::column_spec(
    1,
    width = "4cm",
    extra_css = "white-space: nowrap;"
  ) %>%
  kableExtra::column_spec(
    2:9,
    width = "2.2cm",
    extra_css = "white-space: nowrap; text-align: center;"
  ) %>%
  kableExtra::scroll_box(
    width = "100%"
  )
```

### Sensitivity analysis results table

```{r display-final-results-sensitivity}
# Display the final results table for the sensitivity analysis
knitr::kable(
  final_results_sensitivity,
  caption = "Sensitivity analysis results",
  align = "lcccccccc",
  format = "html"
) %>%
  kableExtra::kable_styling(
    full_width = TRUE,
    position = "center",
    font_size = 13
  ) %>%
  kableExtra::column_spec(
    1,
    width = "4cm",
    extra_css = "white-space: nowrap;"
  ) %>%
  kableExtra::column_spec(
    2:9,
    width = "2.2cm",
    extra_css = "white-space: nowrap; text-align: center;"
  ) %>%
  kableExtra::scroll_box(
    width = "100%"
  )
```

```{r export-final-tables}
# Define the output folder for final results tables
final_tables_dir <- "results/final_tables"

# Create the folder if it does not already exist
dir.create(
  final_tables_dir,
  recursive = TRUE,
  showWarnings = FALSE
)

# Export the final results tables as CSV files
readr::write_csv(
  final_results_main,
  file.path(final_tables_dir, "main_analysis_results.csv")
)

readr::write_csv(
  final_results_sensitivity,
  file.path(final_tables_dir, "sensitivity_analysis_results.csv")
)
```
The final results tables were exported to the `results/final_tables` folder as: 
- `main_analysis_results.csv`
- `sensitivity_analysis_results.csv`

These files contain the final statistical results for the main analysis and the sensitivity analysis.

### Final results table interpretation

The main analysis included **461 participants**. Both wrist extension and forearm supination showed statistically significant positive changes between T0 and T1.

- For wrist extension, the mean value increased from **50.21 degrees** at T0 to **51.89 degrees** at T1. The mean delta was **1.68 degrees** and the median delta was **1.70 degrees**. The Wilcoxon estimated shift was **1.70 degrees**, with a 95% confidence interval from **1.60 to 1.85 degrees**. **The result was statistically significant (P < 0.001)**;

- For forearm supination, the mean value increased from **69.70 degrees** at T0 to **71.74 degrees** at T1. The mean delta was **2.04 degrees** and the median delta was **2.10 degrees**. The Wilcoxon estimated shift was **2.10 degrees**, with a 95% confidence interval from **1.90 to 2.25 degrees**. **The result was statistically significant (P < 0.001)**.

The sensitivity analysis excluded the **7 participants** flagged during the anatomical plausibility check and included **454 participants**. The results remained statistically significant for both outcomes.

- For wrist extension, the mean delta was **1.69 degrees** and the Wilcoxon estimated shift was **1.70 degrees**, with a 95% confidence interval from **1.60 to 1.85 degrees**;
- For forearm supination, the mean delta was **2.03 degrees** and the Wilcoxon estimated shift was **2.05 degrees**, with a 95% confidence interval from **1.90 to 2.25 degrees**.

These sensitivity results were very close to the main analysis results. Therefore, the main statistical conclusions were not driven by the anatomically flagged participants.

Overall, both wrist extension and forearm supination showed **statistically significant positive changes between T0 and T1**. However, because this was a pre-post analysis without a control group, these findings should be interpreted as observed within-participant changes rather than definitive causal effects of the prevention program.

## Final figures

A final faceted boxplot figure was used to visualize the distributions of wrist extension and forearm supination at T0 and T1.

This figure complements the final statistical tables by showing the distribution of individual participant values, the median, the interquartile range, and the mean value for each outcome at each timepoint.

```{r final-figure-boxplots, fig.width=9, fig.height=5.5, fig.align="center", fig.cap="Distribution of wrist extension and forearm supination at T0 and T1."}
# Prepare the data for the final figure
final_plot_data <- analysis_df %>%
  select(
    participant_id,
    wrist_extension_deg_T0,
    wrist_extension_deg_T1,
    forearm_supination_deg_T0,
    forearm_supination_deg_T1
  ) %>%
  tidyr::pivot_longer(
    cols = -participant_id,
    names_to = c("Outcome", "Timepoint"),
    names_pattern = "(.*)_deg_(T0|T1)",
    values_to = "Degrees"
  ) %>%
  mutate(
    Outcome = dplyr::recode(
      Outcome,
      "wrist_extension" = "Wrist extension",
      "forearm_supination" = "Forearm supination"
    ),
    Outcome = factor(
      Outcome,
      levels = c("Wrist extension", "Forearm supination")
    ),
    Timepoint = factor(
      Timepoint,
      levels = c("T0", "T1")
    )
  )

# Create the final boxplot figure
final_boxplot <- ggplot(
  final_plot_data,
  aes(x = Timepoint, y = Degrees, fill = Timepoint)
) +
  geom_boxplot(
    alpha = 0.65,
    width = 0.55,
    outlier.shape = NA,
    color = "black"
  ) +
  geom_jitter(
    aes(color = Timepoint),
    width = 0.12,
    alpha = 0.25,
    size = 1.1
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 23,
    size = 3,
    fill = "white",
    color = "black"
  ) +
  facet_wrap(
    ~ Outcome,
    scales = "free_y"
  ) +
  scale_fill_manual(
    values = c(
      "T0" = "#3B6EA8",
      "T1" = "#C65A1E"
    )
  ) +
  scale_color_manual(
    values = c(
      "T0" = "#3B6EA8",
      "T1" = "#C65A1E"
    )
  ) +
  labs(
    title = "Range-of-motion outcomes at T0 and T1",
    x = "Timepoint",
    y = "Degrees (°)",
    fill = "Timepoint",
    color = "Timepoint"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    strip.text = element_text(face = "bold"),
    legend.position = "bottom"
  )

# Display the figure in the HTML report
final_boxplot
```
```{r export-final-figure}
# Define the output folder for the final figure
final_figures_dir <- "results/final_figures"

# Create the folder if it does not already exist
dir.create(
  final_figures_dir,
  recursive = TRUE,
  showWarnings = FALSE
)

# Export the final boxplot figure as a PNG file
ggsave(
  filename = file.path(final_figures_dir, "final_boxplot_range_of_motion.png"),
  plot = final_boxplot,
  width = 9,
  height = 5.5,
  dpi = 300
)
```
The final boxplot figure was exported to the `results/final_figures` folder as `final_boxplot_range_of_motion.png`

_Final figure interpretation_:

The final figure shows the distributions of wrist extension and forearm supination at T0 and T1.

For both outcomes, the T1 distribution is slightly shifted upward compared with T0. This visual pattern is consistent with the descriptive results and the Wilcoxon signed-rank tests, which showed statistically significant positive changes between T0 and T1.

The individual points show variability between participants. The white diamonds represent the mean values, which were slightly higher at T1 for both wrist extension and forearm supination.

The figure should be interpreted as a descriptive visual summary of the observed pre-post changes. It supports the statistical results visually, but it does not establish a causal effect of the Ostéart prevention program because the study design did not include a control group.

# Discussion

The aim of this project was to describe changes in wrist extension and forearm supination between T0 and T1 in leather workers who participated in the 6-month Ostéart prevention program.

The main analysis showed statistically significant positive changes for both outcomes. Wrist extension increased by a mean of **1.68 degrees**, and forearm supination increased by a mean of **2.04 degrees**. The Wilcoxon estimated shifts were **1.70 degrees** for wrist extension and **2.10 degrees** for forearm supination. These results indicate a positive change between baseline and follow-up for both range-of-motion outcomes.

However, the magnitude of the observed changes was modest. This distinction is important because statistical significance does not necessarily imply a large clinical or functional change. With a sample of **461 participants**, even small changes can reach statistical significance. Therefore, the results should be interpreted as statistically robust but modest in size.

The observed range-of-motion values were compared with reference values reported by Norkin and White. For wrist extension, the AAOS reference value is **70 degrees**, while the AMA considers **60 degrees or more** as functional. In the present dataset, mean wrist extension increased from **50.21 degrees at T0** to **51.89 degrees at T1**. Therefore, although the change was statistically significant, the mean values remained below these reference values.

For forearm supination, the AAOS reference value is **80 degrees**, while the AMA considers **80 degrees or more** as functional. In the present dataset, mean forearm supination increased from **69.70 degrees at T0** to **71.74 degrees at T1**. These mean values also remained below the AAOS and AMA reference values. This suggests that the observed increase did not bring the group mean up to usual reference values, but it did indicate a small positive shift.

These comparisons suggest that the observed mean values were anatomically plausible but lower than commonly reported reference values. However, reference values should be interpreted cautiously. Range-of-motion values can vary according to age, sex, limb dominance, measurement method, occupational exposure, pain, fatigue, and assessment conditions. In this project, the participants were leather workers, a population potentially exposed to repetitive hand and wrist movements, which may partly explain lower mean values than those reported in general reference sources.

The sensitivity analysis excluding the **7 participants** flagged during the anatomical plausibility check produced results very close to the main analysis. This suggests that the main statistical conclusions were not driven by these high but plausible values. This also supports the decision to retain them in the main analysis while checking their influence separately.

Direct comparison of the observed deltas with previous intervention studies is limited. I did not identify a directly comparable study involving leather workers, a 6-month prevention program, and the same two range-of-motion outcomes. Therefore, the results should mainly be interpreted as observed within-participant changes in this dataset rather than as effects directly comparable to previous intervention trials.

Finally, because the study used a pre-post design without a control group, the results cannot establish a causal effect of the Ostéart prevention program. They indicate modest observed improvements over time, but other explanations cannot be ruled out, such as measurement variability, natural variation, changes in work exposure, or other contextual factors during the follow-up period.

# Conclusion

This project analysed an initial range-of-motion dataset collected among 473 leather workers who participated in the 6-month Ostéart prevention program. After Python data preparation, the final complete-case analysis included **461 participants.** The statistical analysis was then performed in R.

Both wrist extension and forearm supination showed **statistically significant positive changes between T0 and T1**. These results remained consistent after excluding the 7 participants flagged during the anatomical plausibility check.

Overall, the findings suggest modest observed improvements in the two range-of-motion outcomes between baseline and follow-up. However, because the study used a pre-post design without a control group, these findings should be interpreted as observed within-participant changes rather than definitive causal effects of the Ostéart prevention program.

# References

Norkin, CC et White, DJ (2017) Mesure du mouvement articulaire : un guide de goniométrie. 5e édition, FA Davis Company, Philadelphie.

TM INSTITUTE MONTPELLIER, https://www.tminstitute.fr/osteart.html, Ostéart'Pro©, Visited in 2026.

Keith T. Palmer, E. Clare Harris, David Coggon, Carpal tunnel syndrome and its relation to occupation: a systematic literature review, Occupational Medicine, Volume 57, Issue 1, January 2007, Pages 57–66, https://doi.org/10.1093/occmed/kql125

Nicoletti S, Carino M, Di Leone, et al. G. Prevalence of upper limb work-related musculoskeletal disorders (UL-WMSDs) in workers of the upholstered furniture industry. Med Lav. 2008;99(4):271-280. doi:10.23749/mdl.v99i4.1683
